---
title: "Aphid Ectotherm Model"
format: html
editor: visual
---

```{r}
#| label: setup

library(NicheMapR)
library(dplyr)
library(tidyr)    # pivot_wider() in the last chunk
library(ggplot2)
library(sf)
library(mcera5)
library(here)     # paths resolve from the 3_theory folder (3_theory.Rproj)
```

## Ectotherm Model Parameters for Aphid Species

The below script was initially assembled using Claude - careful review and validation of each parameter is required before any further steps.

```{r}

# =============================================================================
# NicheMapR ectotherm() parameters for three cereal aphids
#   Metopolophium dirhodum, Rhopalosiphum padi, Sitobion avenae
#
# Purpose: body-temperature (Tb) simulations for the Rothamsted trap sites,
# to be post-processed with TDT / TTL damage accrual (heat + cold branches)
# fitted from data/aphid_tdt.csv.
#
# Status of each value is tagged:
#   [DATA]   from your own aphid_tdt.csv (Li et al. 2025 JEB design)
#   [LIT]    taken from a published source (see refs at bottom)
#   [ASSUME] reasoned default - check / measure / run sensitivity
#
# No published NicheMapR heat-budget parameterization exists for these
# species (or any aphid) that I could find. What is reused:
#   - NicheMapR's built-in ellipsoid geometry (shape = 2)
#   - AmP DEB entries for other aphids (optional block at the end)
#   - Your TDT data for thermal limits
# =============================================================================


# -----------------------------------------------------------------------------
# 1. Species-level trait table
# -----------------------------------------------------------------------------
aphid_traits <- tibble::tribble(
  ~species,                  ~code,        ~Ww_g,    ~shape_b, ~shape_c, ~alpha, ~CT_max, ~CT_min, ~T_pref, ~T_F_min, ~T_F_max, ~Usrhyt, ~minshade, ~maxshade,
  "Metopolophium dirhodum",  "M_dirhodum", 0.59e-3,  0.45,     0.35,     0.70,   35.5,    -4.6,    20.0,    5.0,      28.0,     0.15,    30,        90,
  "Rhopalosiphum padi",      "R_padi",     0.41e-3,  0.60,     0.45,     0.80,   36.9,    -3.7,    25.0,    5.0,      28.0,     0.05,    50,        90,
  "Sitobion avenae",         "S_avenae",   0.56e-3,  0.45,     0.35,     0.75,   36.4,    -2.7,    22.5,    5.0,      28.0,     0.50,     0,        80
)
# Column notes
#  Ww_g      [LIT] wet mass of 6-d (3rd instar) aphids, Li et al. 2025 JEB:
#            S. avenae 0.56 mg, M. dirhodum 0.59 mg, R. padi 0.41 mg.
#            Same cohort design as your TDT data, so Tb and tolerance match.
#            Apterous adults are heavier (~1 mg): rerun with adult mass too.
#            At <1 mg, Tb is almost insensitive to mass (Tb ~ local air/leaf).
#  shape_b,  [ASSUME] prolate ellipsoid (shape = 2), long axis = 'a'.
#  shape_c   width:length ~0.45 for spindle-shaped Macrosiphini (S. avenae,
#            M. dirhodum), ~0.6 for broadly oval R. padi. Height:length ~0.35-0.45.
#            With rho = 1000 kg/m3 these imply body lengths of ~1.9 mm
#            (S. avenae, M. dirhodum) and ~1.4 mm (R. padi): plausible for
#            late instars.
#  alpha     [ASSUME] solar absorptivity. Pale green M. dirhodum lowest,
#            olive/dark R. padi highest, S. avenae (green or brown morphs)
#            intermediate. Not measured for these species: priority to measure
#            (spectrometer on live aphids) or bracket 0.6-0.9 in sensitivity.
#  CT_max    [DATA] temperature giving LT50 = 60 min on your heat TDT curve
#            (ages pooled; z = 4.4, 3.6, 3.5 C for Md, Rp, Sa). Hourly-step
#            equivalent of a lethal limit. Used by NicheMapR only for retreat
#            decisions (and death if CT_kill = 1, which we keep OFF).
#  CT_min    [LIT] chill-coma temperature, Alford et al. 2014 PLOS ONE:
#            Md -4.6, Sa -2.7, Rp -3.7 (R. padi n = 1, weak).
#            Cold-TDT 60-min LT50 from your data: Md -8.5, Rp -4.8, Sa -8.2.
#  T_pref    [LIT] temperature of maximum development rate, Dean 1974:
#            Md 20, S. (Macrosiphum) avenae 22.5, Rp 25 C.
#  T_F_min   [LIT] lower development threshold. Cereal-aphid literature
#            gives 0-6 C; 5 C is the standard value (Brabec et al. 2014).
#  T_F_max   [ASSUME] Dean 1974: all nymphs died by 30 C at constant
#            temperature. 28 C used as the upper feeding/performance limit.
#  Usrhyt    [ASSUME] height (m) in the crop for the microclimate run.
#            Classic vertical partitioning in wheat: S. avenae on flag leaf /
#            ear, M. dirhodum on mid-lower leaves, R. padi at stem base.
#            Change with crop stage.
#  minshade, [ASSUME] canopy shade (%). Lower in canopy -> more shade.
#  maxshade  maxshade = refuge on leaf underside / deeper in canopy.

# Fitted TDT parameters from data/aphid_tdt.csv (for post-processing Tb)
# Method: binomial GLM of survival ~ log10(duration) per species x age x temp
# -> LT50; then OLS log10(LT50) ~ temp, ages pooled. Refit properly in your
# TDT pipeline (e.g. with age as a covariate); these are first-pass values.
aphid_tdt_pars <- tibble::tribble(
  ~code,        ~branch, ~z,    ~T_LT50_1min, ~T_LT50_60min, ~T_LT50_600min, ~R2,
  "M_dirhodum", "heat",   4.39,  43.3,         35.5,          31.1,           0.97,
  "R_padi",     "heat",   3.57,  43.2,         36.9,          33.3,           0.96,
  "S_avenae",   "heat",   3.54,  42.7,         36.4,          32.9,           0.97,
  "M_dirhodum", "cold",  -1.53, -11.2,         -8.5,          -6.9,           0.88,
  "R_padi",     "cold",  -2.26,  -8.8,         -4.8,          -2.5,           0.64,
  "S_avenae",   "cold",  -1.66, -11.2,         -8.2,          -6.5,           0.92
)
# Sanity check against Li et al. 2025: LT50 at 34 C for 6-d aphids from these
# data = 116 (Md), 635 (Rp), 368 (Sa) min vs published 125, 696, 375 min.

# -----------------------------------------------------------------------------
# 2. Parameters shared by all three species
# -----------------------------------------------------------------------------
aphid_common <- list(
  # Geometry / body
  shape     = 2,      # ellipsoid
  shape_a   = 1,
  orient    = 0,      # prolate, lying lengthwise on the leaf
  rho_body  = 1000,   # kg/m3 [ASSUME]
  c_body    = 3073,   # J/kg/K, NicheMapR default
  k_flesh   = 0.5,    # W/m/K, NicheMapR default
  epsilon   = 0.95,   # IR emissivity, standard for arthropod cuticle [ASSUME]

  # Radiation / contact
  # NicheMapR's 'substrate' is the SOIL SURFACE. An aphid on a leaf does not
  # touch soil, so conduction is switched off. Half the IR view is to the sky
  # and half downward (canopy/ground); the downward view still sees soil
  # temperature, which is the main structural approximation here.
  pct_cond  = 0,
  pct_touch = 0,
  fatosk    = 0.5,
  fatosb    = 0.5,
  conv_enhance = 1,

  # Behaviour
  live       = 2,  # behaviour on, but no orientation to the sun
  postur     = 3,  # normal silhouette, not tracking the sun
  diurn      = 1,
  nocturn    = 1,  # phloem feeding continues at night
  crepus     = 1,
  shade_seek = 1,  # move to leaf underside / lower canopy
  burrow     = 0,  # no soil retreat
  climb      = 0,
  delta_shade = 3,

  # Heat budget mode
  transient  = 0,  # at ~0.5 mg the thermal time constant is seconds, so
                   # steady state at an hourly step is appropriate
  enberr     = 0.01, # default; the solver tolerance is enberr * mass, so it
                     # already scales down for a sub-mg animal

  # Water loss (minor for Tb at this size; aphids drink phloem continuously)
  pct_wet   = 0.05,  # % surface as free water [ASSUME]
  pct_eyes  = 0,
  pct_mouth = 0,
  pantmax   = 1,

  # Metabolism: V_O2 (ml/h) = M_1 * M^M_2 * 10^(M_3*Tb) * 10^M_4, M in g.
  # Metabolic heat is negligible for Tb at <1 mg; these only affect O2 and
  # respiratory water outputs. M_3 = log10(Q10)/10 with Q10 = 2.
  # M_1 is calibrated below to a placeholder resting rate.
  M_2 = 0.82,  # insect interspecific mass exponent ~0.75-0.82 [ASSUME]
  M_3 = log10(2) / 10,
  M_4 = 0,

  # Death / energetics switched off: TDT damage is handled afterwards
  CT_kill = 0,
  DEB     = 0
)

# Calibrate M_1 so that V_O2 = target at 20 C for a given mass.
# Placeholder target: 3 ul O2 / mg / h at 20 C [ASSUME]. Replace with a
# measured value (e.g. respirometry from the Li et al. 2025 metabolic work).
calib_M1 <- function(Ww_g, vo2_ul_mg_h = 3, Tref = 20, M_2, M_3) {
  vo2_ml_h <- vo2_ul_mg_h * (Ww_g * 1000) / 1000
  vo2_ml_h / (Ww_g^M_2 * 10^(M_3 * Tref))
}
```

## Build Argument Function for `ectotherm()` for each species

```{r}
# -----------------------------------------------------------------------------
# 3. Build the argument list for one species
# -----------------------------------------------------------------------------
aphid_ecto_args <- function(code, micro) {
  tr <- aphid_traits[aphid_traits$code == code, ]
  stopifnot(nrow(tr) == 1)

  # Every microclimate input is passed explicitly. ectotherm()'s defaults
  # otherwise look for an object called `micro` in the global environment,
  # and a stale one there gives a mismatched number of days.
  ndays <- nrow(micro$metout) / 24
  rain  <- micro$RAINFALL
  if (length(rain) != ndays) {        # e.g. monthly totals from micro_global
    rain <- rep(sum(rain) / ndays, ndays)
  }

  micro_inputs <- list(
    nyears    = micro$nyears,
    metout    = micro$metout,    shadmet   = micro$shadmet,
    soil      = micro$soil,      shadsoil  = micro$shadsoil,
    soilmoist = micro$soilmoist, shadmoist = micro$shadmoist,
    humid     = micro$humid,     shadhumid = micro$shadhumid,
    soilpot   = micro$soilpot,   shadpot   = micro$shadpot,
    tcond     = micro$tcond,     shadtcond = micro$shadtcond,
    rainfall  = rain,
    rainhr    = rep(-1, nrow(micro$metout)),
    preshr    = rep(101325 * ((1 - (0.0065 * as.numeric(micro$elev) / 288))^(1 / 0.190284)),
                    nrow(micro$metout)),
    DEP       = micro$DEP,
    KS        = micro$KS[seq(1, 19, 2)],
    b         = micro$BB[seq(1, 19, 2)],
    PE        = abs(micro$PE[seq(1, 19, 2)]) * -1,
    alpha_sub = 1 - micro$REFL,
    elev      = as.numeric(micro$elev),
    longitude = as.numeric(micro$longlat[1]),
    latitude  = as.numeric(micro$longlat[2])
  )

  args <- c(aphid_common, micro_inputs, list(
    Ww_g      = tr$Ww_g,
    shape_b   = tr$shape_b,
    shape_c   = tr$shape_c,
    alpha_max = tr$alpha,
    alpha_min = tr$alpha,
    CT_max    = tr$CT_max,
    CT_min    = tr$CT_min,
    T_pref    = tr$T_pref,
    T_F_min   = tr$T_F_min,
    T_F_max   = tr$T_F_max,
    T_B_min   = tr$T_F_min,  # basking is not meaningful for aphids
    T_RB_min  = tr$T_F_min,
    minshades = rep(tr$minshade, ndays * 24),
    maxshades = rep(tr$maxshade, ndays * 24)
  ))
  args$M_1 <- calib_M1(tr$Ww_g, M_2 = args$M_2, M_3 = args$M_3)
  args
}

```

## Get Microclimate Data

Here, we will extract microclimate data for the Rothamsted site Broom's Barn for 2024. We will need to set seperate bounding boxes for each site to avoid downloading huge .nc files.

```{r}

# -----------------------------------------------------------------------------
# 4. Example run at one trap
# The microclimate must be run separately for each species because Usrhyt
# (height in the crop) and shade differ. The minshade/maxshade in micro must
# match the values passed to ectotherm.

# -----------------------------------------------------------------------------
#get bbox of our trap sites
roth_coords <- read.csv(here("data", "rothamsted_coordinates.csv"))
#run only at one trap for now, broom's barn
bb<-c(roth_coords[1,3],roth_coords[1,2])
sf_bb <- st_as_sf(data.frame(lon = bb[1], lat = bb[2]), coords = c("lon", "lat"), crs = 4326)
bbox <- st_bbox(sf_bb)

#buildmcera5 request
req <- build_era5_request(xmin = bbox[1], xmax = bbox[3], 
                          ymin = bbox[2], ymax = bbox[4],
                          start_time = as.POSIXct("01/01/2024", format = "%d/%m/%Y"),
                          end_time = as.POSIXct("31/12/2024", format = "%d/%m/%Y"), 
                          outfile_name = "era5_roth_2024")

#this takes forever - unsilence when ready
#request_era5(request = req, out_path = here("data", "mcera5"),overwrite=F)
```

## Run Aphid Ecotherm Model for One site

Here, we run the aphid microclimate model for Broom's Barn site throughout 2024.

```{r}
#| label: run-aphid
#| warning: false
#| cache: true

run_aphid <- function(code, loc, dstart,dfinish) {
  tr <- aphid_traits[aphid_traits$code == code, ]
  micro <- micro_era5(loc = loc, Usrhyt = tr$Usrhyt,
                        minshade = tr$minshade, maxshade = tr$maxshade,
                        dstart = dstart, dfinish = dfinish,spatial = here("data", "mcera5", "era5_roth_2024"),
                        run.gads = 2)  # R version of the aerosol model; the Fortran one (default) can crash a non-interactive R session
  args  <- aphid_ecto_args(code, micro)
  ecto  <- do.call(ectotherm, args)
  environ <- as.data.frame(ecto$environ)
  if (nrow(environ) == 0) {
    stop("ectotherm() returned no output for ", tr$species,
         ": check the 'error:' messages printed above")
  }
  environ$species <- tr$species
  environ
}



# Rothamsted (Harpenden) trap, lon/lat from data/rothamsted_coordinates.csv
out <- purrr::map_dfr(aphid_traits$code, run_aphid, loc = bb,dstart = "01/01/2024", dfinish = "31/12/2024")
```

We can then plot the resulting body temperatures predicted by the model

```{r}
#| echo: FALSE
#| warning: FALSE
out$date<-as.Date(out$DOY-1,origin="2024-01-01")  # 2024 is a leap year
out$datetime <- as.POSIXct(paste(out$date,out$TIME),format="%Y-%m-%d %H",tz="UTC")
out$TC_anom <- out%>%group_by(datetime,species)%>%mutate(TC_anom=TC-TA)%>%pull(TC_anom)

# Key columns: TC (body temp), SHADE, ACT, DEP (should stay 0: no burrowing)
ggplot() +geom_line(data=out,aes(x=datetime,y=TC,group=species)) +
  facet_wrap(.~species) +
  geom_line(data=out,aes(x=datetime,y=TA,color="Air Temperature, species-level height"),col="red")
```

We can also look at anomalies compared to the ambient air temperature

```{r}
#| echo: FALSE
#| warning: FALSE
#body temperature-air temperature anomalies
ggplot(data = out) +
  geom_line(aes(x = datetime, y = TC_anom, color = species)) +
  labs(title = "Body temperature anomalies (Tb - Ta) for aphid species", x = "Date", y = "Temperature anomaly (C)") +
  theme_minimal() +facet_grid(.~species)
```

As well as anomalies in body temperature between species. Note that species-level differences are due to slightly different parameterizations of the aphid model - we may decide to collapse into a single aphid model.

```{r}
#| echo: FALSE
#| warning: FALSE
#dataframes of anomalies between each pairwise species
anomalys <- out %>% group_by(datetime,species) %>% summarise(TC=mean(TC)) %>% pivot_wider(names_from = species, values_from = TC) %>% 
  mutate(Md_Rp=`Metopolophium dirhodum`-`Rhopalosiphum padi`,Md_Sa=`Metopolophium dirhodum`-`Sitobion avenae`,Rp_Sa=`Rhopalosiphum padi`-`Sitobion avenae`)

ggplot() +
  geom_line(data=anomalys,aes(x=datetime,y=Md_Rp)) +
  geom_line(data=anomalys,aes(x=datetime,y=Md_Sa),color="red") +
  geom_line(data=anomalys,aes(x=datetime,y=Rp_Sa),color="blue") +
  labs(title="Body temperature anomalies between aphid species",x="Date",y="Body temperature anomaly (C)") +
  theme_minimal()
```
