---
title: "Aphid TLS"
format: html
editor: visual
bibliography: references.bib
---

```{r}
#| setup: true
#| message: false
#| echo: false


library(tidyverse)
library(bayesTLS) #requires brms, Rcpp, STAN installation. I use rstan backend on windows, but you should use cmdstanr
library(ggplot2)
library(here)
library(rstan)
library(brms)
library(lubridate)
library(ggpubr)

```

## Cereal Aphid TLS

In their recent preprint, @noble used time to death data from @li2023 to construct Bayesian thermal tolerance landscapes using the `bayesTLS` package. Conveniently, the three species in the paper (*Metopolophium dirhodum*, *Sitobion avenae* and *Rhopalosiphum padi*) are the same three species we extracted abundnace data for from 15 sites in the Rothamsted insect survey (`rothamsted_data_readin.qmd`). For future use, we replicate their analysis (found at <https://daniel1noble.github.io/bayesTLS/#sec-aphids>) here.

## Data Read-in

```{r}
#| warning: false
#| message: false
#their package includes the data we need - how convenient!
data(aphid_tdt)
aphid_raw <- aphid_tdt |>
  dplyr::filter(branch == "heat", age == "6") |>
  dplyr::mutate(species = droplevels(species))

aphid_std <- standardize_data(aphid_raw, temp = "temp", duration = "duration_min",
                              n_total = "n_total", n_surv = "n_surv",
                              duration_unit = "minutes")   # keeps `species`
str(aphid_std)
```

## Construct 4PL TLS Model

We use the Noble et al. code here to construct a 4PL TLS model. Note that

```{r}
#| warning: false
#| message: false
# Species effects on z and CTmax with a temperature
# slope on k, but we constrain the asymptotes to be constant (up = ~ 1, low = ~ 1).
# Letting up/low vary by species x temperature overfits here and there is littl evidence up 
# and low vary dramatically with temperature.
system.time(
wf_aphid <- fit_4pl(aphid_std,
                    ctmax = ~ 0 + species,
                    z = ~ 0 + species,
                    k = ~ temp_c,
                    up = ~ 1,
                    low = ~ 1,
                    t_ref = 60, silent = 2, refresh = 0,
                    backend    = "rstan",  #delete line if using cmdstanR
                    file_refit = "on_change",
                    file       = here("data", "aphid_species")))
summary(wf_aphid)
```

## Per-species TDT parameters

We can extract canonical TDT parameters for each species.

```{r}
#|echo: false
#|warning: false

aphid_tls <- tls(wf_aphid, by = "species", lethal = TRUE)
aphid_est <- get_tls_est(aphid_tls)

# One row per species: median [95% CI] for each quantity.
aphid_tls_tbl <- aphid_est |>
  dplyr::mutate(
    value    = format_interval(median, lower, upper, digits = 1),
    quantity = factor(quantity, levels = c("z", "CTmax", "Tcrit"),
                      labels = c("z (°C)", "CTmax 1h (°C)", "Tcrit (°C)")),
    species  = dplyr::recode(species, M_dirhodum = "M. dirhodum",
                             R_padi = "R. padi", S_avenae = "S. avenae")) |>
  dplyr::select(species, quantity, value) |>
  tidyr::pivot_wider(names_from = quantity, values_from = value) |>
  dplyr::rename(Species = species)
tinytable::tt(aphid_tls_tbl)

```

## Thermal Tolerance Landscapes

We can plot the survival curves across experimental temperatures

```{r}
#| warning: false
#| echo: false

psc_aphid <- predict_survival_curves(wf_aphid)
# Show the full fitted curves (clip_to_observed = FALSE): each species' complete
# sigmoid across the shared duration range, rather than capping each temperature
# at its longest tested exposure.
plot_survival_curves(psc_aphid, observed = aphid_std, log_time = TRUE,
                     clip_to_observed = FALSE)

```

## Moving forward to predicting injury and survival

Refer to the [vignette](https://daniel1noble.github.io/bayesTLS/#sec-aphids) to see how they predict survival over temperature traces. We would be able to accomplish something similar with microclimates/measured temperatures. Functions to use are `predict_heat_injury()` and `plot_heat_injury()`.

Below is an example for one summer at one site. Note that the `predict_heat_injury` function takes a while to run - 800 seconds on my machine.

```{r}
#| echo: false
#| warning: false
#| cache: true
micro_flat <- readRDS(here("data", "micro_uk", "micro_uk_flat.rds"))

broom_2019 <- micro_flat %>%
  filter(Trap == "Brooms_Barn_S",
         datetime >= as.POSIXct("2019-06-01", tz = "UTC"),
         datetime <  as.POSIXct("2019-09-01", tz = "UTC")) %>%
  transmute(time = (datetime-min(datetime))/dhours(1), hod = hour(datetime), temp = TALOC)

str(broom_2019)
system.time(
aphid_death_brooms_2019<-predict_heat_injury(broom_2019,
                    wf_aphid, target_surv = "relative",
                    trace_unit = "hours", by = "species"))
ggarrange(nrow=2,
ggarrange(ncol=3,
          ggplot(data = broom_2019,aes(x=time,y=temp))+geom_line() +theme_minimal(),
          ggplot(data = broom_2019,aes(x=time,y=temp))+geom_line()+theme_minimal(),
          ggplot(data = broom_2019,aes(x=time,y=temp))+geom_line()+theme_minimal()),
plot_heat_injury(aphid_death_brooms_2019))
```
